Designing Types for R, Empirically

### 2020-11-01 paper/main.tex
first part edits

R's main data type is the primitive vector.  Vectors care explicitly
constructed by the constructor \code{c(...)}, as in \code{c(1L, 2L,
  3L)} which creates a vector of three integers.  R has a builtin
notion of type that can be queried by the \k{typeof}
function. Figure~\ref{rtypes} lists all of the builtin types that are
provided by the language; these are the possible return values of
\k{typeof}. There is no intrinsic notion of subtyping in R, but in
many contexts a \k{logical} will be coerced to \k{integer}, and an
\k{integer} will be coerced to \k{double}.  Some odd conversions can
occur in corner cases, such as \k{1<"2"} holds and \k{c(1,2)[1.6]}
returns the first element of the vector, as the double is converted to
an integer. R does not distinguish between scalars and vectors (they
are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
  typeof(c(5,5))} \code{ == "double"}.  With the exception of lists,
all vectorized data types are monomorphic.  A \k{list} can hold values
of any other type including \k{list}. For all monomorphic data types,
attempting to store a value of a different type will cause a
conversion. Either the value is converted to the type of the vector,
or the vector is converted to the type of the value.

@@

\noindent
All vectorized data types have a distinguished missing value denoted
by \code{NA} (for ``not available'').  The type of \code{NA} is
\k{logical} (\code{typeof(NA)=="logical"}), but \k{NA} inhabits every
type: \code{typeof(c(1,NA)[2])} \k{==} \k{"double"}.  R also has a
\code{NULL} value.  In data science, it is useful to have a notion of
a ``missing observation'', since vectors are monomorphic there is a
need for a missing value to inhabit each primitive type. In a sense,
\code{NA} represents a missing data point, and \code{NULL} is a
missing data set.

Over time, programmers have found the need for a richer type
structure.  R supports this with {\it attributes}. One may think of
attributes as an optional map from names to values attached to any
built in type.  Attributes are used to encode various type
structures. They are queried with \k{attributes} and \k{class}.  Using
attributes, programmers can extend the set of types by tagging
data. For example, take the vector of four values,
\code{x<-c(1,2,3,4)} and attach attribute \k{dim},
\code{attr(x,"dim")<-c(2,2)}, to treat \k{x} as a 2x2 matrix.

Another attribute that can be set is the \k{class}, this is one way to
use R for object-oriented programming.  This attribute can be bound to
a list of class names. For instance, \code{class(x)<-"human"} will set
the class of \k{x} to be \k{human}.  There are three
object-orientation frameworks in R: S3, S4, and R5.  The S3 object
system support single dispatch on the class of the first argument of a
function, whereas the S4 object system allows multiple dispatch. R5
allows for users to define objects in a more imperative style. Some of
the most widely used data types leverage attributes, e.g., data frames
and matrices. A data frame, for instance, is a list of vectors with a
class and a column name attribute, and matrices are vectors with a
 \k{dims} attribute.

@@

\noindent
This function has four formal parameters, \k x, dots, \k y and \k
z. Parameter \k x can be bound positionally or passed by name. The
dots, \k{...}, is always positional. The remaining two parameters must
be passed by name as they are preceded by dots; \k y and \k z have
default values, in the case of \k z this is a constant, but \k{y} is
bound to an expression that depends on \k z's value (if \k z is not
zero, \k y defaults to \k{NULL}). The body of the function will add \k
x and \k y tp either 0 or the result of concatenating the dots into a
primitive vector. The function \k{missing} tests if a parameter was
explicitly passed. The following are some valid invocations of \k f:

@@

Given the pecularities of the language there are a number of design
choices that need to be reviewed and evaluated. It is not
controversial to include types that cover the six kinds of primitive
vectors, furthermore environments and the distinguished null types are
commonly used and must be included.  Environments are lists with
reference semantics: mutating a value in an environment is performed
in-place.  They are used to store variables and to escape from the
copy-on-write semantics of other data types.  For simplicity, we omit
some of the data types that closer to the implementation of the
language. Fig.~\ref{types} presents our type language.

@@

\paragraph{Scalar} While R does not have scalar data types, there are
cases where functions expect scalar values, for example a conditional
takes a single logical (and will complain if more values are passed).
We considered tracking dimensions of data structures, but decided
against it. Instead, the type language differentiates between vectors
of lenght 1 and vectors of any dimension.  The primitive types can be
either vectors (e.g., \VEC\int) or scalars (e.g., \int).  A vector can
happen to be of length 1, and thus a scalar is also a vector.  Vectors
are monomorphic, a vector of doubles contains only doubles.

\paragraph{Missing} 
Each of basic types has its specific \NA. Many built-in functions,
especially those implemented in C or Fortran, do not support \NA
values.  It is thus advantageous to distinguish between vector that
can contain missing values and those that are guaranteed not to.  In
our experience, functions that expect scalar values tend to not admit
{\NA}s, thus scalar types are treated as being \NA-free. The type
language does allows to write \VEC\int to specify that a vector of
integers may have {\NA}s and \NAVEC\int to say that a vector must not
have missing values. Of course, a value of type \VEC\int may happen to
not have any \NA and thus be of type \NAVEC\int.  The type \raw does
not allow {\NA}s, so \VEC\raw is \NA-free.

\paragraph{Nullable} 
The \nul type is inhabited by a singleton \code{NULL} value often used
as a sentinel. Unlike in some other lanugage, \k{NULL} is not the
default value of uninitialized variables. R has different notion for
that (which we do not cover here).  To capture common uses, of
\k{NULL}, the type language has a nullable type, written \code{?}\T.
Values of this type can be either values of type \T or \k{NULL}.

\paragraph{Lists} Heterogeneous collections are implemented using lists.
Lists and vectors are closely related: a vector converts to a list
with \code{as.list}, and lists to vectors with \code{unlist}
(coercions may ensue). The type language allows to specify that a
value is a list containing element of some type \T, written \LIST\T. R
does not have built-in type tests for this purpose, to establish the
type of a value requires traversal and checking individual elements.

\paragraph{Class}
R has three objects systems, code-named S3, S4 and R5. All of them
operate by adding a class attribute to values. The most widely used
system, by far, is S3 which supports single dispatch and multiple
classes~\cite{ecoop12}. The challenge from a type system point of view
is that a value such as integer 5, could be attributed with a class.
Code that performs dispatch would use the class attribute while code
that does not would view the value as an integer. The type language
focuses on the attribute and will hide the underlying type of the
value. While that seems to match common usage it does represent a loss
of expressiveness. The type language also focuses on S3, we leave the
other object systems for future work. S3 has no notion of inheritance,
each value has a list of classes. The type language thus allows to
write \CLASS{\ID_1,\dots,\ID_n} to denote values tagged with the class
names $\ID_1$ to $\ID_n$.

\paragraph{Union} We support untagged unions of types written
$\T_1 \k{~|~} ... \k{~|~} \T_n$. The elements of unions are not
disjoint, 1L is both an \int, a \VEC\int and a \NAVEC\int. R does not
provide a built-in type testing mechanism, e.g. for the case of
\NA-free data types, it is necessary to scan vectors to find out if
they have missing values.

\paragraph{Function}
Functions signatures the form \FUN{A_1, \dots, A_n}\T where each
$\A_i$ argument is either a type \T or \VARGS, a variable length
argument list. Moreover, a single function's signature can be the
disjunction of a number of individual signatures.

@@

The choice of type language follows the structure of values.  The
presence of \NA-free data types and scalars are two choices that must
be validated in practice.  Nullable types are just a special case of
unions.

While R does not support the notion of subtyping between values.  The
conversions between primitive types give us a starting point (e.g., an
\int is always accepted where a \dbl is expected). Furthermore, the
types introduced above induce some the rules of Figure~\ref{fig:sub}:
a vector without \NA is a subtype of a vector, a value of type \T is a
subtype of a nullable \T, a list is subtype of another list if their
elements are subtypes, and a scalar is a subtype of a vector of the
same primitive type.

\begin{figure}[h!tb]\begin{tabular}{rclll}
  \NAVEC\S  & <:& \VEC\S \\
  \T        & <:& \k{?}\T \\
  \LIST{\T} & <:& \LIST{\T'} & \it{iff} \T <: \T' \\
  \S        & <:& \VEC\S \\
  \lgl      & <:& \int\\
  \int      & <:& \dbl\\
  \dbl      & <:& \clx\\  
\end{tabular}
\caption{Subtyping}\label{fig:sub}
\end{figure}

>>>>

R's main data type is the primitive vector.  Vectors are explicitly
constructed by the constructor \code{c(...)}, as in \code{c(1L, 2L,
  3L)} which creates a vector of three integers.  R has a builtin
notion of type that can be queried with the \k{typeof}
function. Figure~\ref{rtypes} lists all of the builtin types
provided by the language; these are the possible return values of
\k{typeof}. There is no intrinsic notion of subtyping in R, but in
many contexts a \k{logical} will be coerced to \k{integer}, an
\k{integer} will be coerced to \k{double}, and a \k{double} will be coerced to \k{complex}.  
Some odd conversions occur in corner cases, such as \k{1<"2"} holds and \k{c(1,2)[1.6]}
returns the first element of the vector, as the double is converted to
an integer. R does not distinguish between scalars and vectors (they
are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
  typeof(c(5,5))} \code{ == "double"}. All vectorized data types are monomorphic, except \k{list} which can hold values
of any type, including \k{list}. For all monomorphic data types,
attempting to store a differently-typed value will cause a
conversion: Either the value is converted to the type of the vector, or vice versa.

@@

All vectorized data types have a distinguished missing value denoted
by \code{NA} (for ``not available'').  The type of \code{NA} is
\k{logical} (\code{typeof(NA)=="logical"}), but \k{NA} inhabits every
type: \code{typeof(c(1,NA)[2])} \k{==} \k{"double"}.  R also has a
\code{NULL} value.  In data science, it is useful to have a notion of
a ``missing observation'', since vectors are monomorphic there is a
need for a missing value to inhabit each primitive type. In a sense,
\code{NA} represents a missing data point, whereas \code{NULL} represents a
missing data set.

Over time, programmers have found the need for a richer type
structure.  R supports this with {\it attributes}. One may think of
attributes as an optional map from names to values attached to any
built in type.  Attributes are used to encode various type
structures, and they are queried with \k{attributes}.  Using
attributes, programmers can extend the set of types by tagging
data. For example, take the vector of four values,
\code{x<-c(1,2,3,4)} and attach attribute \k{dim},
\code{attr(x,"dim")<-c(2,2)}, to treat \k{x} as a 2x2 matrix.

Another attribute that can be set is the \k{class}, which is one way to
use R for object-oriented programming. The class attribute can be queried with \k{class}, and it can be bound to
a list of class names: For instance, \code{class(x)<-c("human", "friend")} will set
the class of \k{x} to be \k{human} and \k{friend}.  There are three
object-orientation frameworks in R: S3, S4, and R5.  The S3 object
system support single dispatch on the class of the first argument of a
function, whereas the S4 object system allows multiple dispatch. R5
allows for users to define objects in a more imperative style. Some of
the most widely used data types leverage attributes, e.g., data frames
and matrices. A data frame, for instance, is a list of vectors with a
class and a column name attribute, and matrices are vectors with a
 \k{dims} attribute.

@@

\noindent
This function has four formal parameters, \k x, dots, \k y and \k
z. Parameter \k x can be bound positionally or passed by name. The
dots, \k{...}, is always positional. The remaining two parameters must
be passed by name as they are preceded by dots; \k y and \k z have
default values, in the case of \k z this is a constant, but \k{y} is
bound to an expression that depends on \k z's value (if \k z is not
zero, \k y defaults to \k{NULL}). The body of the function will add \k
x and \k y to either 0 or the result of concatenating the dots into a
primitive vector. The function \k{missing} tests if a parameter was
explicitly passed. The following are some valid invocations of \k f:

@@

Given the peculiarities of the language there are a number of design
choices that need to be reviewed and evaluated. It is not
controversial to include types that cover the six kinds of primitive
vectors, furthermore environments and the distinguished null types are
commonly used and must be included.  Environments are lists with
reference semantics: mutating a value in an environment is performed
in-place.  They are used to store variables and to escape from the
copy-on-write semantics of other data types.  For simplicity, we omit
some of the data types that closer to the implementation of the
language. Fig.~\ref{types} presents our type language.

@@

\paragraph{Scalar} While R does not have scalar data types, there are
cases where functions expect scalar values, for example a conditional
takes a single logical and will complain if more values are passed.
We considered tracking dimensions of data structures, but decided
against it. Instead, the type language differentiates between vectors
of lenght 1 and vectors of any dimension.  The primitive types can be
either vectors (e.g., \VEC\int) or scalars (e.g., \int).  A vector can
happen to be of length 1, which we handle with subtyping, shown later.  Vectors
are monomorphic, a vector of doubles contains only doubles.

\paragraph{Missing} 
Each primitive type has its specific \NA. Many built-in functions,
especially those implemented in C or Fortran, do not support \NA
values.  It is thus advantageous to distinguish between vectors that
can and cannot contain missing values.  In
our experience, functions that expect scalar values tend to not admit
{\NA}s, thus scalar types are treated as being \NA-free. The type
language allows one to write \NAVEC\int to specify that a vector of
integers {\it may} have {\NA}s and \VEC\int to say that a vector must not
have missing values.  The type \raw does
not allow {\NA}s, so \NAVEC{\raw} is \NA-free.
R does not provide a built-in type testing mechanism, e.g., for the case of
\NA-free data types, it is necessary to scan vectors to find out if
they have missing values.

\paragraph{Nullable} 
The \nul type is inhabited by a singleton \code{NULL} value often used
as a sentinel. Unlike in some other lanugage, \k{NULL} is not the
default value of uninitialized variables. R has different notion for
that (which we do not cover here).  To capture common uses of
\k{NULL}, the type language has a nullable type, written \code{?}\T.
Values of this type can be either values of type \T or \k{NULL}.

\paragraph{Lists} Heterogeneous collections are implemented using lists.
Lists and vectors are closely related: a vector converts to a list
with \code{as.list}, and lists to vectors with \code{unlist}
(coercions may ensue). The type language allows one to specify that a
value is a list containing element of some type \T, written \LIST\T. R
does not have built-in type tests for this purpose, to establish the
type of a value requires traversal and checking individual elements.

\paragraph{Class}
R has three objects systems, code-named S3, S4 and R5. All of them
operate by adding a class attribute to values. The most widely used
system by far is S3, which supports single dispatch and multiple
classes~\cite{ecoop12}. The challenge from a type system point of view
is that a value (such as an integer \k{5L}) could be attributed with a class.
Code that performs dispatch would use the class attribute while code
that does not would view the value as an integer. The type language
focuses on the attribute, and will hide the underlying type of the
value. While this seems to match common usage, it does represent a loss
of expressiveness. The type language also focuses on S3, and we leave the
other object systems for future work. S3 has no notion of inheritance,
each value has a list of classes. The type language thus allows to
write \CLASS{\ID_1,\dots,\ID_n} to denote values tagged with the class
names $\ID_1$ to $\ID_n$.

\paragraph{Union} We support untagged unions of types written
$\T_1 \k{~|~} ... \k{~|~} \T_n$. The elements of unions are not
disjoint, e.g., 1L is both an \int, a \VEC\int and a \NAVEC\int.

\paragraph{Function}
Functions signatures the form \FUN{A_1, \dots, A_n}\T where each
$\A_i$ argument is either a type \T or dots (\VARGS), a variable length
argument list. Moreover, a single function's signature can be the
disjunction of a number of individual signatures.

Overall, the choice of type language follows the structure of values.  The
presence of \NA-free data types and scalars are two choices that must
be validated in practice.  Note that nullable types are really just a special case of
unions (i.e., \code{?}\T is shorthand for $\T \k{~|~} \k{NULL}$).

@@

R does not support the notion of subtyping between values, but we include a few simple rules in
our framework to capture widespread coercion.  The
conversions between primitive types give us a starting point (e.g., an
\int is always accepted where a \dbl is expected). Furthermore, the
types introduced above induce some the rules of Figure~\ref{fig:sub}:
an \NA-free vector is a subtype of its \NA-ful equivalent, a value of type \T is a
subtype of a nullable \T, a list is subtype of another list if their
elements are subtypes, and a scalar is a subtype of a vector of the
same primitive type.

\begin{figure}[h!tb]\begin{tabular}{rclll}
  \VEC\S    & <:& \NAVEC\S \\
  \T        & <:& \k{?}\T \\
  \LIST{\T} & <:& \LIST{\T'} & \it{iff} \T <: \T' \\
  \S        & <:& \VEC\S \\
  \lgl      & <:& \int\\
  \int      & <:& \dbl\\
  \dbl      & <:& \clx\\  
\end{tabular}
\caption{Subtyping}\label{fig:sub}
\end{figure}
